Missing the Point\(\colon\) Non-Convergence in Iterative Imputation Procedures

### 2020-04-16 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
Editing the methods section to be conform recommendations Morris et al.

Negative $AC$-values indicate divergence within imputation chains. **Subsequent sampled values within each imputation chain are less alike. High $AC$-values are implausible in MI procedures. That is, the randomness induced by the MI algorithm effectively mitigates the risk of dependency within chains.**

@@

Based on an empirical finding [@lace07], we hypothesize that $\widehat{R}$ will over-estimate non-convergence of MI algorithms. The threshold of $\widehat{R} < 1.01$ will then be too stringent for diagnosing convergence. This over-estimation may, however, be diminished because $\widehat{R}$ can falsely diagnose convergence if initial values of the algorithm are not appropriately over-dispersed [@broo98, p. 437]. In `mice`, initial values are chosen randomly from the observed data. Therefore, we cannot be certain that the initial values are over-dispersed. We expect this to have little effect on the hypothesized performance of $\widehat{R}$. **(Add actual hypothesis: we thought that mice would converge sooner than $\widehat{R}$ would indicate that it did)** No hypothesis was formulated about the performance of $AC$ as convergence diagnostic.

@@

The convergence of `mice` is investigated through model-based simulation in `R` [version `r paste(version$major,version$minor, sep = ".")`; @R]. The simulation set-up is summarized in the pseudo-code below. The complete `R` script of the simulation study is available from [github.com/gerkovink/shinyMice](https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy).

<!-- Simulation conditions are the number of iterations ($T = 1, 2, \dots, 100$) and the percentage missingness in the  -->

@@

```{r, eval=FALSE, echo=TRUE}
# pseudo-code of simulation 
simulate data 
for (number of simulation runs from 1 to 1000)
  for (number of iterations from 1 to 100)
    create missingness
    impute the missingness
    compute convergence diagnostics
    perform analysis
    pool results
    compute simulation diagnostics
aggregate convergence and simulation diagnostics
```

**(Add step for different missingness proportions.)**

A finite population of $N=1000$ is simulated to solve a multiple linear regression problem, where dependent variable $Y$ is regressed on independent variables $X_1$, $X_2$ and $X_3$ (**check notation with betas here, and add what the quantity/-ies of scientific interest is/are!**):

@@

\begin{align*}
\begin{pmatrix}X_1\\
X_2\\
X_3\\
\epsilon
\end{pmatrix} \sim  N
\begin{bmatrix}
\begin{pmatrix}
12\\
3\\
0.5\\
0
\end{pmatrix}\!\!,
\begin{pmatrix}
4 & 4 & 1.8 & 0\\
4 & 16 & 4.8 & 0\\
1.8 & 4.8 & 9 & 0\\
0 & 0 & 0 & 100
\end{pmatrix}
\end{bmatrix}\\[2\jot]
\end{align*}

>>>>

Negative $AC$-values indicate divergence within imputation chains. **Subsequent sampled values within each imputation chain are less alike.**

@@

Based on an empirical finding [@lace07], we hypothesize that $\widehat{R}$ will over-estimate non-convergence of MI algorithms. The threshold of $\widehat{R} < 1.01$ will then be too stringent for diagnosing convergence. This over-estimation may, however, be diminished because $\widehat{R}$ can falsely diagnose convergence if initial values of the algorithm are not appropriately over-dispersed [@broo98, p. 437]. In `mice`, initial values are chosen randomly from the observed data. Therefore, we cannot be certain that the initial values are over-dispersed. We expect this to have little effect on the hypothesized performance of $\widehat{R}$. **(Add actual hypothesis: we thought that mice would converge sooner than $\widehat{R}$ would indicate that it did)** No hypothesis was formulated about the performance of $AC$ as convergence diagnostic. **Added later: High $AC$-values are implausible in MI procedures. That is, the randomness induced by the MI algorithm effectively mitigates the risk of dependency within chains.**

@@

Convergence of the MICE algorithm is investigated through model-based simulation in `R` [version `r paste(version$major,version$minor, sep = ".")`; @R]. The simulation set-up is summarized in the pseudo-code below. The complete `R` script of the simulation study is available from [github.com/gerkovink/shinyMice](https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy). 
<!-- Simulation conditions are the number of iterations ($T = 1, 2, \dots, 100$) and the percentage missingness in the  -->

@@

```
# pseudo-code of simulation 
1. simulate data 
for (missingness proportions 5%, 25%, 50%, 75% and 95%)
  for (number of simulation runs from 1 to 1000)
    2. create missingness
    for (number of iterations from 1 to 100)
      3. impute missingness
      4. compute convergence diagnostics
      5. perform analysis of scientific interest
      6. pool results across imputations
      7. compute performance measures
  8. aggregate outcomes 
9. combine outcomes
```
The aim of the simulation study is to evaluate the impact of: 1) terminating the MICE algorithm at different imputation chain lengths ($t = 1, 2, ..., 100$), and 2) varying the difficulty of the missingness problem (**different missingness proportions** $prop = .05, .25, .50, .75, .95$). For the purpose of this study, sampling variance is not of interest. Therefore, a single complete set may serve as comparative truth in all simulation runs [@vinknd].

The data-generating mechanism is a multivariate normal distribution, representing person data on three independent variables (from an unspecified social scientific field of study). A finite population of $N=1000$ is simulated using the function `mvtnorm::rmvnorm()`, where

\begin{align*}
\begin{pmatrix}X_1\\
X_2\\
X_3\\
\epsilon
\end{pmatrix} \sim  N
\begin{bmatrix}
\begin{pmatrix}
12\\
3\\
0.5\\
0
\end{pmatrix}\!\!,
\begin{pmatrix}
4 & 4 & 1.8 & 0\\
4 & 16 & 4.8 & 0\\
1.8 & 4.8 & 9 & 0\\
0 & 0 & 0 & 100
\end{pmatrix}
\end{bmatrix}\!\!\text{, and}\\[2\jot]
Y &=  1 + 2X_1 + .5X_2 - X_3 + \epsilon .
\end{align*} 
Subsequently, 
We consider four quantities of scientific interest ['conceptual estimands'; @morr19], namely  The quantity of scientific interest is the estimated regression coeffiecent estimand is the A finite population of $N=1000$ is simulated to solve a multiple linear regression problem, where dependent variable $Y$ is regressed on independent variables $X_1$, $X_2$ and $X_3$ (**check notation with betas here, and add what the quantity/-ies of scientific interest is/are!**):

@@

### 2020-04-16 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Editing the methods section to be conform recommendations Morris et al.

Negative \(AC\)-values indicate divergence within imputation chains.
\textbf{Subsequent sampled values within each imputation chain are less
alike. High \(AC\)-values are implausible in MI procedures. That is, the
randomness induced by the MI algorithm effectively mitigates the risk of
dependency within chains.}

@@

Based on an empirical finding \citep{lace07}, we hypothesize that
\(\widehat{R}\) will over-estimate non-convergence of MI algorithms. The
threshold of \(\widehat{R} < 1.01\) will then be too stringent for
diagnosing convergence. This over-estimation may, however, be diminished
because \(\widehat{R}\) can falsely diagnose convergence if initial
values of the algorithm are not appropriately over-dispersed
\citep[p.~437]{broo98}. In \texttt{mice}, initial values are chosen
randomly from the observed data. Therefore, we cannot be certain that
the initial values are over-dispersed. We expect this to have little
effect on the hypothesized performance of \(\widehat{R}\). \textbf{(Add
actual hypothesis: we thought that mice would converge sooner than
\(\widehat{R}\) would indicate that it did)} No hypothesis was
formulated about the performance of \(AC\) as convergence diagnostic.

@@

The convergence of \texttt{mice} is investigated through model-based
simulation in \texttt{R} \citep[version 3.6.3;][]{R}. The simulation
set-up is summarized in the pseudo-code below. The complete \texttt{R}
script of the simulation study is available from
\href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy}{github.com/gerkovink/shinyMice}.

\begin{verbatim}
# pseudo-code of simulation 
simulate data 
for (number of simulation runs from 1 to 1000)
  for (number of iterations from 1 to 100)
    create missingness
    impute the missingness
    compute convergence diagnostics
    perform analysis
    pool results
    compute simulation diagnostics
aggregate convergence and simulation diagnostics
\end{verbatim}

\textbf{(Add step for different missingness proportions.)}

@@

@@

\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-4-1} \end{flushleft}

\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-4-2} \end{flushleft}

\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-4-3} \end{flushleft}

>>>>

Negative \(AC\)-values indicate divergence within imputation chains.
\textbf{Subsequent sampled values within each imputation chain are less
alike.}

@@

Based on an empirical finding \citep{lace07}, we hypothesize that
\(\widehat{R}\) will over-estimate non-convergence of MI algorithms. The
threshold of \(\widehat{R} < 1.01\) will then be too stringent for
diagnosing convergence. This over-estimation may, however, be diminished
because \(\widehat{R}\) can falsely diagnose convergence if initial
values of the algorithm are not appropriately over-dispersed
\citep[p.~437]{broo98}. In \texttt{mice}, initial values are chosen
randomly from the observed data. Therefore, we cannot be certain that
the initial values are over-dispersed. We expect this to have little
effect on the hypothesized performance of \(\widehat{R}\). \textbf{(Add
actual hypothesis: we thought that mice would converge sooner than
\(\widehat{R}\) would indicate that it did)} No hypothesis was
formulated about the performance of \(AC\) as convergence diagnostic.
\textbf{Added later: High \(AC\)-values are implausible in MI
procedures. That is, the randomness induced by the MI algorithm
effectively mitigates the risk of dependency within chains.}

@@

Convergence of the MICE algorithm is investigated through model-based
simulation in \texttt{R} \citep[version 3.6.3;][]{R}. The simulation
set-up is summarized in the pseudo-code below. The complete \texttt{R}
script of the simulation study is available from
\href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy}{github.com/gerkovink/shinyMice}.

\begin{verbatim}
# pseudo-code of simulation 
1. simulate data 
for (missingness proportions 5%, 25%, 50%, 75% and 95%)
  for (number of simulation runs from 1 to 1000)
    2. create missingness
    for (number of iterations from 1 to 100)
      3. impute missingness
      4. compute convergence diagnostics
      5. perform analysis of scientific interest
      6. pool results across imputations
      7. compute performance measures
  8. aggregate outcomes 
9. combine outcomes
\end{verbatim}

The aim of the simulation study is to evaluate the impact of: 1)
terminating the MICE algorithm at different imputation chain lengths
(\(t = 1, 2, ..., 100\)), and 2) varying the difficulty of the
missingness problem (\textbf{different missingness proportions}
\(prop = .05, .25, .50, .75, .95\)). For the purpose of this study,
sampling variance is not of interest. Therefore, a single complete set
may serve as comparative truth in all simulation runs \citep{vinknd}.

The data-generating mechanism is a multivariate normal distribution. We
simulate a finite population of \(N=1000\) to represent person data for
an unspecified social scientific field of study.

\begin{align*}
\begin{pmatrix}X_1\\
X_2\\
X_3\\
\epsilon
\end{pmatrix} &\sim  N
\begin{bmatrix}
\begin{pmatrix}
12\\
3\\
0.5\\
0
\end{pmatrix}\!\!,
\begin{pmatrix}
4 & 4 & 1.8 & 0\\
4 & 16 & 4.8 & 0\\
1.8 & 4.8 & 9 & 0\\
0 & 0 & 0 & 100
\end{pmatrix}
\end{bmatrix}\!\!\text{, and}\\[2\jot]
Y &=  1 + 2X_1 + .5X_2 - X_3 + \epsilon .
\end{align*}

@@

\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-2-1} \end{flushleft}

\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-2-2} \end{flushleft}

\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-2-3} \end{flushleft}

@@
